New results in $\sfrac{k}{n}$ Power-Hours

### 2013-12-28 powerhour/paper/paper.tex
Writing paper

\newcommand\comment[1]{}
\newcommand\sfrac[2]{{}\,^{#1}\!/{}\!_{#2}}

@@

\renewcommand\>{$>$}
\newcommand\<{$<$}
\newcommand\kn{\ensuremath{\sfrac{k}{n}\,}}

\newcommand\any{\ensuremath{\textrm{?}}}
\newcommand\nocup{\text{\sout{\ensuremath{\cup}}}}
\newcommand\fullcup{\ensuremath{\uplus}}
\newcommand\emptycup{\ensuremath{\cup}}
\newcommand\overcup{\ensuremath{\cap}}

@@

% XXX TODO: Display in 3D cube.
It is also feasible overnight to enumerate all three-player games. The
results are three-dimensional, of course, but we can display the
outcomes for two of the players in the familiar presentation
(Figure~\ref{fig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
achievable for any $k_1 \in \{ 0 .. 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
and some unknown $k_3$ (projected out of this display). This is a
significant improvement over what was achievable in BMML with two
players. It suggests that with enough friends willing to follow a
program, some set of people are likely to be able to achieve any
amount of drinks between them (if they have the ability to construct
the right rule set!); see Section~\ref{sec:conjectures}.

@@

@@

Like any drinking game, there are several arbitrary things about BMML.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{soloncups.pdf}
\end{center}\vspace{-0.1in}
\caption{Possible outcomes for BM$n$L with a single player. With no
  cup states, it is only possible to drink nothing. By convention, the
  0\th\ cup is the special ``filled'' state, so it is only possible to
  drink on every turn (60) or never (0). As we add more cup states,
  the number of achievable states strictly increases; with 8 cup states
  we can drink any amount. 7 and 53 drinks are the most difficult to
  achieve, and can only be done with 8 cup states.
}
\label{fig:soloncups}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
\end{center}\vspace{-0.1in}
\caption{What's achievable for the first two players in a three-player
  generalized BM$n$L game, with $n$ ranging from 1 cup state (darkest)
  to 5 (lightest). TODO WRITEME
% 6 cupstates ran overnight and got here:
% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
}
\label{fig:composite2player}
\end{figure}

@@

Define BMnL, where $n$ is the number of non-filled states. BMML=BM2L.
(XXX actually I think it's the total number of states including filled
everywhere but that sentence, right?)  Find $n$ such that a solo or
dual power hour achieves any $k$.

\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L.
  \item With two players, we know that BM5L (XXX check) does not allow
    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
    can achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1
    \}$. No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}).
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, based on how many samples we need to get close to 100\%
    for BM7L.
  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
    possible, given how close BM7L is for a single player. We know it
    can get 99.79\%, at least!
  \item With three players one player can drink any number of shots in
    BMML with two other players helping.
\end{enumerate}

@@

Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_n \rangle$ is possible,
so is $\langle k_1, \ldots, 0, \ldots, k_n \rangle$ (under what conditions)?

>>>>

\newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

@@

\renewcommand\>{$>$}
\newcommand\<{$<$}
\newcommand\kn{\ensuremath{\sfrac{k}{n}\,}}
\newcommand\bmsl{BM$s$L}

\newcommand\any{\ensuremath{\textrm{?}}}
\newcommand\nocup{\text{\sout{\ensuremath{\cup}}}}
\newcommand\fullcup{\ensuremath{\uplus}}
\newcommand\emptycup{\ensuremath{\cup}}
\newcommand\overcup{\ensuremath{\cap}}
\newcommand\scup[1]{\ensuremath{\cup\!\!\!{}^{{}_{#1}}}}

@@

% XXX TODO: Display in 3D cube.
It is also feasible overnight to enumerate all three-player games. The
results are three-dimensional, of course, but we can display the
outcomes for two of the players in the familiar presentation
(Figure~\ref{fig:3and2}). Note that $\langle k_1, k_2, k_3 \rangle$ is
achievable for any $k_1 \in \{ 0 \ldots 60 \}$, $k_2 \in \{ 0, 1, 2 \}$,
and some unknown $k_3$ (projected out of this display). This is a
significant improvement over what was achievable in BMML with two
players. It suggests that with enough friends willing to follow a
program, some set of people are likely to be able to achieve any
amount of drinks between them (if they have the ability to construct
the right rule set!); see Section~\ref{sec:conjectures}.

@@

Like any drinking game, there are several arbitrary things about BMML.
While we will not tamper with its essence (for example, allowing beer
to be spilled from a cup without drinking it), there are some other
variables to adjust. The most naturally flexible is the number of cup
states. We will always have \fullcup, a cup with beer in it. In BMML
we also have \emptycup\ and \overcup. But why not $\supset$ (cup
turned on its side, facing west) or $\stackrel{\wedge}{\cup}$ (an
upright cup with a cocktail umbrella on it)?

We define \bmsl, where $s$ is the number of distinct cup states. By
convention, the 0\th\ cup state will be the filled cup \fullcup\ since
it has special rules. The remainder will be \scup{i} for $i \in \{ 1,
\ldots, s - 1 \}$. BMML is BM3L where we've just renamed \emptycup\ to
\scup{1}, and \overcup\ to \scup{2}.

Clearly, more cups give us more expressive power, and should allow us
to reach more outcomes. To illustrate, recall the construction of
$k=31$ in the solo BM3L game (Figure~\ref{fig:solo31}). It has a
length-2 cycle alternating between two cup states, where the player
drinks on every other turn. The third state is just used once as a
drinking lead-in to make the total 31 rather than 30. With an
additional cup state, we can straightforwardly transform this into a
game with an outcome of $k=32$ by extending the prelude with another
state where the player drinks. Of course, the expressive power is not
just limited to such extensions; we now can create cycles of new
lengths, admit the possibility of more disconnected cycles, and so on.

It is easy to enumerate the 1-player \bmsl\ games. These appear in
Figure~\ref{fig:soloncups}. The interesting range for $s$ is $\{ 1
\ldots 8 \}$.\!\footnote{It's not clear that BM0L should be considered
  legal as the rules speak of a 0\th\ cup, but it is degenerate
  anyway.} At 8 cup states, the player can achieve any amount in a
solo game. Importantly, this extends to any number of players in BM8L,
because the players can pass to themselves and not even interact.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{soloncups.pdf}
\end{center}\vspace{-0.1in}
\caption{Possible outcomes for \bmsl\ with a single player. The
  vertical axis shows an increasing number of cup states, and the
  horizontal axis shows the achievable values of $k$ drinks. With no
  cup states, it is only possible to drink nothing. By convention, the
  0\th\ cup is the special ``filled'' state, so it is only possible to
  drink on every turn (60) or never (0). As we add more cup states,
  the number of achievable states strictly increases; with 8 cup
  states we can drink any amount. 7 and 53 drinks are the most
  elusive, and can only be done with 8 cup states.}
\label{fig:soloncups}
\end{figure}

The two-player case is much more interesting. We've already enumerated
all the possible outcomes for BM3L
(Figure~\ref{fig:powerhour2player}). It is computationally tractable
to enumerate them for $s < 6$. The set of achievable outcomes in
BM$s$L is always contained within BM$(s+1)$L, because we can embed a
game from the former into the latter by just never producing the
additional cup state and having any arbitrary rule for it. Therefore,
we show these results in a composite grid where more and more states
are reachable as we increase $s$ (Figure~\ref{fig:composite2player}).

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{composite2playermock.png}
\end{center}\vspace{-0.1in}
\caption{What's achievable for two players in a generalized \bmsl\ game,
  with $s$ ranging from 1 cup state (darkest) to 5 (lightest).
% 6 cupstates ran overnight and got here:
% #26186200000 queue 109 min 49 466729 states/sec 796272 config/sec
}
\label{fig:composite2player}
\end{figure}

In order to efficiently enumerate BM5L, I improved the algorithm
again. Observe that the ``drink'' action associated with a rule
usually does not affect anything but the final outcome. The only
exception is that in the rule for \fullcup, the player must drink if
passing in a state other than \fullcup. Putting this aside for a
moment, note that we can just count the number of times each rule was
executed for each player, producing an $s$-dimensional vector $[ d_1,
  \ldots d_s ]$ for each player. That player is able to achieve many
different drink totals, specifically, $d_1 \times r_1 + \ldots + d_s
\times r_s$ where $r_i$ is 1 if the player should drink on that rule
and 0 if not. Simulating a game this way is even more like abstract
interpretation (we leave a concrete value free and compute a formula
rather than an integer), and allows us to evaluate many concrete games
at once. At the end, we simply plug in every legal value for $r_i$ for
each player and insert those games into the database. This last step
is where we must tend to the exception around \fullcup. We may not set
$r_0$ to 0 if the player ever passes \fullcup\ in a non-full state. A
very close approximation would be to insist that $r_0 = 1$ if the rule
in the \fullcup\ position does not output as \fullcup, but this is
inexact, as that rule may never be executed\footnote{We may be able to
  argue that in that case, there always exists another game that does
  not violate this condition. But I think it is simpler to just
  implement the rules.} Instead, during simulation we keep track of
whether each player ever actually passed a non-full cup from the
\fullcup\ state. If so, then we force $r_0 = 1$, which attends to this
special case.

Although this makes earlier enumerations extremely fast and BM5L
quite quick, 2-player BM6L ran 26 billion concrete states overnight
and made only modest progress. In the absence of fancier techniques
for reducing the state space, we must resort to different, inexact
approaches.

\section{The fractal geometry of \kn\ Power-Hours}

Note that all of the two-dimensional figures resemble one another even
though they are fundamentally different (adding players, adding
states, adding random trials). This suggests that the combinatorial
problem (``what outcomes are reachable from finite state machines that
look kind of like this?'') has some geometric structure.

@@

@@

Define \bmsl, where $s$ is the number of cup states. BMML=BM3L. Find
$s$ such that a solo or dual power hour achieves any $k$.

\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L.
  \item With two players, we know that BM5L (XXX check) does not allow
    all combinations of $\langle k_1, k_2 \rangle$ shots, though BM5L
    can achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1
    \}$. No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}).
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, given that random exploration plateaus at about 95.54\%.
  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
    possible, given how close BM7L is for a single player. We know it
    can get 99.89\%, at least!
  \item With three players one player can drink any number of shots in
    BMML with two other players helping.
\end{enumerate}

@@

Teetotaller: When $\langle k_1, \ldots, k_i, \ldots, k_p \rangle$ is possible,
so is $\langle k_1, \ldots, 0, \ldots, k_p \rangle$ (under what conditions)?
